---
title: "EDLD Final Project"
author: "Lauren Berny, Adriana Conn, Mavis Gallo, Aubrey Green, Shawn McWeeney"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(rio)
library(readr)
library(haven)
library(here)
library(tidyverse)
library(janitor)
library(plotly)
library(ggridges)
library(kableExtra)
library(party)

```

# Passengers Data

## Load data
```{r, include = FALSE}
dat <- import(here("data", "dat.csv")) %>% 
   filter(Category=="Passenger") %>% 
   clean_names() %>% 
   mutate_all(na_if,"") %>% 
   drop_na(survived, gender, class, age)
```

## Clean data
```{r, include = FALSE}
dat$gender <- as.factor(dat$gender)
dat$marital_status <- as.factor(dat$marital_status)
dat$category <- as.factor(dat$category)
dat$class <- as.factor(dat$class)
dat$survived <- as.factor(dat$survived)
dat$embarked <- as.factor(dat$embarked)
dat$disembarked  <- as.factor(dat$disembarked)
dat$boat <- as.factor(dat$boat)


dat <- dat %>% 
 mutate(nationality2 = case_when(nationality == "English" ~ "English",
   nationality == "Irish" ~ "Irish",
   nationality == "American" ~ "American",
   nationality == "Swedish" ~ "Swedish",
   nationality == "Finnish" ~ "Finnish",
   nationality == "Scottish" ~ "Scottish",
   nationality == "French" ~ "French",
   nationality == "Italian" ~ "Italian",
   nationality == "Canadian" ~ "Canadian",
   nationality == "Bulgarian" ~ "Bulgarian",
   nationality == "Croatian" ~ "Croatian",
   nationality == "Belgian" ~ "Belgian",
   nationality == "Norwegian" ~ "Norwegian",
   nationality == "Channel Islander" ~ "Channel Islander",
   nationality == "Welsh" ~ "Welsh",
   nationality == "Swiss" ~ "Swiss",
   nationality == "German" ~ "German",
   nationality == "Danish" ~ "Danish",
   nationality == "Spanish" ~ "Spanish",
   nationality == "Australian" ~ "Australian",
   nationality == "Polish" ~ "Polish",
   nationality == "South African" ~ "South African",
   nationality == "Bosnian" ~ "Bosnian",
   nationality == "Hong Kongese" ~ "Hong Kongese",
   nationality == "Dutch" ~ "Dutch",
   nationality == "Lithuanian" ~ "Lithuanian",
   nationality == "Greek" ~ "Greek",
   nationality == "Portuguese" ~ "Portuguese",
   nationality == "Uruguayan" ~ "Uruguayan",
   nationality == "Chinese" ~ "Chinese",
   nationality == "Slovenian" ~ "Slovenian",
   nationality == "Cape Verdean" ~ "Cape Verdean",
   nationality == "Egyptian" ~ "Egyptian",
   nationality == "Japanese" ~ "Japanese",
   nationality == "Hungarian" ~ "Hungarian",
   nationality == "Bosnian" ~ "Bosnian",
   nationality == "Hong Kongese" ~ "Hong Kongese",
   nationality == "Latvian" ~ "Latvian",
   nationality == "Austrian" ~ "Austrian",
   nationality == "Greek" ~ "Greek",
   nationality == "Mexican" ~ "Mexican",
   nationality == "Sweden" ~ "Swedish",
   nationality == "Turkish" ~ "Turkish",
   nationality == "Slovenian" ~ "Slovenian",
   nationality == "Guyanese" ~ "Guyanese",
   nationality == "Haitian" ~ "Haitian",
   nationality == "Syrian,Lebanese" ~ "Syrian/Lebanese",
   nationality == "Unknown" ~ "Unknown",
   TRUE ~ "Other - Multiple", ))


dat <- dat %>% 
   mutate(nationality2 = ifelse(nationality2 == "Unknown", NA, nationality2))
```

## Descriptives
```{r, descriptives, results = "asis", include = FALSE}
# Breakdown of passengers by class
dat %>% 
   group_by(class) %>% 
   summarize(count = n()) %>%
   mutate(percent = (count/sum(count))*100) %>% 
   adorn_totals() %>%
   kable(caption = "Breakdown of Passengers by Class",
       col.names = c("Class", "Count", "Percent"),
       digits = 2,
       booktabs = TRUE) %>%
 kable_classic(full_width = F, html_font = "Cambria")

   
# Breakdown of passengers by class and gender
dat %>% 
   group_by(class, gender) %>% 
   summarize(count = n()) %>% 
   mutate(percent = (count/sum(count))*100) %>%
   kable(caption = "Breakdown of Passengers by Class and Gender",
       col.names = c("Class", "Gender", "Count", "Percent"),
       digits = 2,
       booktabs = TRUE) %>%
 kable_classic(full_width = F, html_font = "Cambria")

# Breakdown of passenger nationalities
dat %>% 
   filter(!is.na(nationality2)) %>% 
   group_by(nationality2) %>% 
   summarize(count = n()) %>% 
   mutate(percent = (count/sum(count))*100) %>% 
   arrange(desc(percent)) %>%
   kable(caption = "Breakdown of Passenger Nationalities",
       col.names = c("Nationality", "Count", "Percent"),
       digits = 2,
       booktabs = TRUE) %>%
   kable_styling(fixed_thead = T, full_width = F, html_font = "Cambria", bootstrap_options = c("striped", "hover"))

dat %>%
   filter(!is.na(nationality2)) %>% 
   group_by(nationality_cat) %>% 
   summarize(count = n()) %>% 
   mutate(percent = (count/sum(count))*100) %>% 
   arrange((percent)) %>%
   ggplot(aes(y = nationality_cat)) + 
   geom_col(aes(x = percent), fill = "dark red") + 
   geom_text(aes(x = percent, label=round(percent,2)), hjust = -.2, size = 3) +
   theme_minimal() + 
   labs(x = "Percentage of all passenges", y = "Nationality")

# Breakdown of passenger nationalities by class (all)
dat %>% 
   filter(!is.na(nationality2)) %>% 
   group_by(class, nationality2) %>% 
   summarize(count = n()) %>% 
   mutate(percent = (count/sum(count))*100) %>% 
   arrange(class, desc(percent)) %>%
   kable(caption = "Breakdown of Passenger Nationalities by Class (All)",
       col.names = c("Class", "Nationality", "Count", "Percent"),
       digits = 2,
       booktabs = TRUE) %>%
   kable_styling(fixed_thead = T, full_width = F, html_font = "Cambria", bootstrap_options = c("striped", "hover"))

dat %>% 
   filter(category == "Passenger") %>% 
   filter(!is.na(nationality2)) %>% 
   group_by(class, nationality_cat) %>%
   summarize(count = n()) %>% 
   mutate(percent = (count/sum(count))*100) %>% 
   arrange(class, desc(percent)) %>%
   ggplot(aes(y = nationality_cat)) + 
   geom_col(aes(x = percent), fill = "dark red") + 
   facet_wrap(~fct_infreq(class)) +
   geom_text(aes(x = percent, label=round(percent,2)), hjust = -.1, size = 3) +
   theme_minimal() +
   labs(x = "Percent of passengers by class", y = "Nationality")


# Average age by class
dat %>% 
   group_by(class) %>% 
   summarize(avg_age = mean(age), std_age = sd(age), min_age = min(age), 
             max_age = max(age)) %>%
   kable(caption = "Average Age by Class",
       col.names = c("Class", "Average Age", "SD Age", "Min. Age", "Max. Age"),
       digits = 2,
       booktabs = TRUE) %>%
   kable_classic(full_width = F, html_font = "Cambria")



```

# Examining Survival

## Survival rates
```{r}
# Survival rate by class
dat %>% 
   group_by(class, survived) %>% 
   summarize(count = n()) %>% 
   mutate(percent = (count/sum(count))*100) %>% 
   arrange(class, survived) %>%
   kable(caption = "Survival Rate by Class",
       col.names = c("Class", "Survived", "Count", "Percent"),
       digits = 2,
       booktabs = TRUE) %>%
   kable_classic(full_width = F, html_font = "Cambria")

# Survival rate by gender
dat %>% 
   group_by(gender, survived) %>% 
   summarize(count = n()) %>% 
   mutate(percent = (count/sum(count))*100) %>% 
   arrange(gender, survived) %>%
   kable(caption = "Survival Rate by Gender",
       col.names = c("Gender", "Survived", "Count", "Percent"),
       digits = 2,
       booktabs = TRUE) %>%
   kable_classic(full_width = F, html_font = "Cambria")

# Survival rate by class and gender
dat %>% 
   group_by(class, gender, survived) %>% 
   summarize(count = n()) %>% 
   mutate(percent = (count/sum(count))*100) %>% 
   arrange(class, gender) %>%
   kable(caption = "Survival Rate by Class and Gender",
       col.names = c("Class", "Gender", "Survived", "Count", "Percent"),
       digits = 2,
       booktabs = TRUE) %>%
   kable_classic(full_width = F, html_font = "Cambria")
```

### Density ridges
```{r}

surv_classhist <- dat %>% 
   ggplot(aes(age, class)) +
   geom_density_ridges(aes(fill = factor(survived))) +
   labs(title = "Age Distribution of Survival Status By Class", 
   x = "Age Distribution", y = "Passenger Class") +
   theme_minimal() +
   theme(plot.title = element_text(hjust = 0.5))

surv_classhist + scale_fill_manual(name = "Survival", values = c("blue","grey"))

surv_agehist <- dat %>% 
   ggplot(aes(age, gender)) +
   geom_density_ridges(aes(fill = factor(survived))) +
   labs(title = "Age Distribution of Survival Status By Gender", 
   x = "Age Distribution", y = "Passenger Gender") +
   theme_minimal() +
   theme(plot.title = element_text(hjust = 0.5))

surv_agehist + scale_fill_manual(name = "Survival", values = c("blue","grey"))

surv_ageclass_hist <- dat %>% 
   ggplot(aes(age, gender)) +
   facet_wrap(~class, nrow=3) +
   geom_density_ridges(aes(fill = factor(survived))) +
   labs(title = "Age Distribution of Survival Status By Class and Gender", 
   x = "Age Distribution", y = "Passenger Gender") +
   theme_minimal() +
   theme(plot.title = element_text(hjust = 0.5))

surv_ageclass_hist + scale_fill_manual(name = "Survival", values = c("blue","grey"))

```

## Predicting survival
```{r}
modeldat <- dat %>% 
   select(survived, gender, class, age)

ctree <- ctree(survived ~ gender + class + age, data=modeldat)
```

```{r, fig.width=12, fig.height=9.25}
plot(ctree)

```


# Tickets Prices

## Load data
```{r}
fares <- import(here("data", "avgfare.csv")) %>% 
   clean_names()

fares$accommodation <- as.factor(fares$accommodation)
fares$accommodation <- factor(fares$accommodation, levels = c("First-class parlor suite", "First-class cabin", "Second-class cabin", "Third-class cabin"))
```

## Calculate inflation
```{r}

p1921 <- (17.9/9.7)
fares$fare_1921 <- p1921*fares$fare_1912
fares$fare_1921 <- round(fares$fare_1921, 2)

p1931 <- (15.2/9.7)
fares$fare_1931 <- p1931*fares$fare_1912
fares$fare_1931 <- round(fares$fare_1931, 2)

p1941 <- (14.7/9.7)
fares$fare_1941 <- p1941*fares$fare_1912
fares$fare_1941 <- round(fares$fare_1941, 2)

p1951 <- (26.0/9.7)
fares$fare_1951 <- p1951*fares$fare_1912
fares$fare_1951 <- round(fares$fare_1951, 2)

p1961 <- (29.9/9.7)
fares$fare_1961 <- p1961*fares$fare_1912
fares$fare_1961 <- round(fares$fare_1961, 2)

p1971 <- (40.5/9.7)
fares$fare_1971 <- p1971*fares$fare_1912
fares$fare_1971 <- round(fares$fare_1971, 2)

p1981 <- (90.9/9.7)
fares$fare_1981 <- p1981*fares$fare_1912
fares$fare_1981 <- round(fares$fare_1981, 2)

p1991 <- (136.2/9.7)
fares$fare_1991 <- p1991*fares$fare_1912
fares$fare_1991 <- round(fares$fare_1991, 2)

p2001 <- (177.1/9.7)
fares$fare_2001 <- p2001*fares$fare_1912
fares$fare_2001 <- round(fares$fare_2001, 2)

p2011 <- (224.9/9.7)
fares$fare_2011 <- p2011*fares$fare_1912
fares$fare_2011 <- round(fares$fare_2011, 2)

p2021 <- (274.3/9.7)
fares$fare_2021 <- p2021*fares$fare_1912
fares$fare_2021 <- round(fares$fare_2021, 2)
```

## Reshape data
```{r}
fares_tidy <- fares %>%
   pivot_longer(cols = starts_with("fare"),
   names_to = "year",
   names_prefix = "fare_",
   values_to = "price", names_transform = list(year = as.integer))

fares_tidy_wide <- fares_tidy %>%
   pivot_wider(names_from = "accommodation",
               values_from = "price")

```



## Inflation-adjustment plot
```{r fig.width=8, fig.height=8}

fare_graph <- fares_tidy %>% 
   ggplot(aes(year, price, colour=accommodation)) +
   geom_line() +
   geom_point() +
   scale_colour_brewer(palette="Spectral") +
   facet_wrap(~ accommodation, 4, scales = "free") +
   xlim(1912,2021) +
   theme(panel.spacing = unit(1, "lines")) +
   labs(y = "Price ($USD)", x = "Year", title = "Inflation-Adjusted Titanic Ticket Prices",
        subtitle = "From 1912 to 2021", colour = "Accommodation") +
   theme_minimal()

ggplotly(fare_graph)

```

```{r echo=FALSE, include=FALSE}
firstavg1912 <- "$150.00"
firstavg2021 <- "$4,241.74"
```
When taking inflation rates into consideration, we see that the average price for a first class cabin in 1912 was **`r firstavg1912`**, which today would be **`r firstavg2021`**

For this project, we used several packages. These included Rio  @riopackage, Readr @readrpackage, Here @herepackage, Tidyverse @tidyversepackage, Janitor @janitorpackage, Plotly @plotlypackage, Kableextra @kableExtrapackage, Ggridges @ggridgespackage, and Party @partypackage

# **References**